---
title: "BIOS 611 Project 1"
author: "Christina Zhou"
date: "10/7/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
#project 1
#set working directory
setwd("~/Desktop/611 Notes/bios611-projects-fall-2019-cwzhou/project_1/results")
```

#### Introduction

Homelessness is defined as any person living without a home, typically living on the streets. In North Carolina, the Continuums of Care to the U.S. Department of Housing and Urban Development (HUD) reported approximately 9,268 people experience homelessness on any given day as of January 2018 (1). Shelters and volunteers attempt to lower this number by providing resources that can potentially help a person or family find the means to obtain a home, and thus stop living on the streets. One of these shelters in Durham, North Carolina, is Urban Ministries of Durham (UMD). Established in 1983, UMD is a non-profit organization that focuses on lowering the community's rate of homelessness by providing food, shelter and other resources to those in need. In 2001, three organizations (Durham Community Shelter for HOPE, St. Philip’s Community Cafe, and the United Methodist Mission Society) merged to create UMD as it exists today. UMD strives to holistically lower Durham's homelessness rates with the help from the Durham community and over 4,000 volunteers. UMD provides "food, shelter, and a future" to over 6,000 men, women, and children every year (2). In our exploratory analysis for Project 1, we seek to explore the trends of household types that visit UMD, regarding number of visits and amount of food. For this project, we focus our data analysis on examining the relationship of number of visits and average pounds of food per visit between family households and single households. We would expect family households to obtain more food, and perhaps have less visits to UMD (maybe because they are trying harder to find a sustainable home due to more members at stake). 

#### Data source
```{r urban, echo=FALSE}
#load dataset
urban <- read.delim("~/Desktop/611 Notes/bios611-projects-fall-2019-cwzhou/project_1/data/UMD_Services_Provided_20190719.tsv", header=T)
#change date format
urban$Date <- as.Date(urban$Date, "%m/%d/%Y")
#extract year variable
urban$Year <- format(as.Date(urban$Date, format="%d/%m/%Y"),"%Y")
#View(arrange(urban,Date))
```
The data for our analysis is from UMD. UMD has been collecting data about people who utilize its resources for decades. This data includes an identifier (ID), date of visit, number of people the food from that visit provided for, pounds of food taken from food pantry, amount of clothing taken from clothing closet, and more. The total number of observations of this data is `r nrow(urban)`. Variable descriptions for the UMD data can be found here: https://github.com/biodatascience/datasci611/blob/gh-pages/data/project1_2019/UMD_Services_Provided_metadata_20190719.tsv

#### Data cleaning and preliminary examinations

Since UMD as it is known today was created in 2001, we truncate our dataset to be collected data from January 1, 2001 to the last date of a full year, which is December 31, 2018. Since the data for 2019 is not complete, we omit data that was collected after December 31, 2018.

To examine the trends of visits and food obtained between family and single households, we define an observation as part of a family household if the number of people the food given that visit was provided for more than 1 person, and as a single household otherwise. Because of this, we want to limit our population to those who are not missing data for the Food.Provided.for (defined as number of people the food given that visit was provided for) and Food.Pounds (defined as the number of pounds given for that visit) variables.

```{r edit, echo=FALSE}
#create dataset df1 that remove dates from FUTURE
df1 <- filter(urban, urban$Date <= Sys.Date()) 
#create a dataset df2 that only keeps dates of interest and drops rows with missing data from food variables
df2 = df1 %>%
  arrange(Client.File.Number,Date) %>% 
  select(Date, Client.File.Number,Food.Provided.for, Food.Pounds,Year) %>%
  filter(Date >= "2001-01-01")%>% #to stay up to date with current trends, we will only examine data in the last 10 years (2001-01-01 to 2018-12-31)
  filter(Date <= "2018-12-31") %>%
  drop_na(Food.Provided.for, Food.Pounds)  #only keep rows that do not have missing variables food.provided.for and food.pounds
  
#create dataset df3 that contains summaries per ID: number of visits, maximum number of family members over all visits, minimum number of family members over all visits, mean pounds of food per visit.
df3 = df2 %>%
  group_by(Client.File.Number) %>%
  summarise(Number_of_Visits = n(),
            max_food = max(Food.Provided.for),
            min_food = min(Food.Provided.for),
            mean_food_pounds_per_day = mean(Food.Pounds))
#create table that keeps subjects between 2001-2018 without missing data from food variables, sorted by households, counting number of visits, first year of visit, last year of visit, range between first and last visit.

#table of number of visits, first visit year, last visit year, and range of visits per ID
df2 %>%
  group_by(Client.File.Number)  %>%
  summarise(Number_of_Visits = n(),
            First_Visit_Year = as.numeric(first(Year)),
            Last_Visit_Year = as.numeric(last(Year)),
            Range = Last_Visit_Year-First_Visit_Year + 1) %>%
  arrange(desc(Number_of_Visits), desc(Range), First_Visit_Year) %>%
  head(10) %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

The table above summarizes the number of visits, number of years of visits, and first and last year of each visit for the 10 households with the largest total number of visits to UMD. This table has been sorted in descending order by range of visited years, total number of visits, and the first year of visit.

Since we are primarily interested in the relationship between household type for this analysis, we subset our analysis population further by only keeping observations with the same number of family members for every visit. We do this because we want to keep consistency in our dataset in terms of defining family vs. single households. We note that changing this subset may be done in a different analysis if a later question of interest involves examining changes number of family household over time. Additionally, we remove observations with who do not take food from UMD (meaning, the maximum pounds of food for all visits is 0) since we are only examining subjects who take food. In the first table below are the top 5 remaining observations, sorted in descending order by the number of family members. The second table below are the top 15 remaining observations, sorted in descending order by mean pounds of food per day per household.

```{r remove outliers, echo=FALSE}
df4 = df3 %>% filter(max_food==min_food) %>% #subset those with same number of family members (1+) each visit
    mutate(Number_of_Members = max_food) %>%
    select(Client.File.Number, Number_of_Visits, Number_of_Members, mean_food_pounds_per_day)
  
df4 %>% arrange(desc(Number_of_Members), desc(mean_food_pounds_per_day)) %>%
  head(5)  %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = F)

df4 %>%
  arrange(desc(mean_food_pounds_per_day),desc(Number_of_Members)) %>%
  head(15)  %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

After examining our population, we see that there is one observation with 650 family members, one with 200 family members, and one with 115 family members. Since these are both extreme outliers, and likely mistakes, we remove these observations from our analysis. Likewise, there appears to be typos regarding average pounds of food per household per day. Thus, we remove observations with more than an average of 100 pounds of food per day, since there was a big jump from less than 100 lbs to over 150 lbs (with few observations). In the table below, we display the top 10 households sorted by average pounds of food from the food pantry per day in descending order. For example, the household with the largest average pounds of food per day is household 15022 with about 86 lbs, with a total of 7 visits and 4 total members in the family.

```{r test1, echo=FALSE}
test1 = df4 %>% 
  filter(Number_of_Members >0) %>% #remove observations who do not take food from UMD since we are only examining those who take food
  filter(Number_of_Members <= 100) %>% #remove the observation with 200 family members (likely a typo)
  filter(mean_food_pounds_per_day <= 100) %>%#remove observations with more than an average of 500 pounds of food per day
  select(Client.File.Number,Number_of_Visits,Number_of_Members,mean_food_pounds_per_day)
test1$family = as.factor(ifelse(test1$Number_of_Members>1,1,0)) #create indicator: if food.provided.for > 1 then family = 1 else family = 0
#define family as more than 1 person, as assumed by "provided.food.for"
#7423 observations

#View(test1)
head(arrange(test1, desc(mean_food_pounds_per_day), desc(Number_of_Visits),desc(Number_of_Members)),5) %>%
    kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

Thus, our final analysis population is `r nrow(test1)` households. From the table below, we can see that the split between number of family households and number of single households is relatively even. Recall the variable family is an indicator where family = 1 means the household is a family household and family = 0 means the household is a single household.

```{r household split, echo=FALSE}
test1 %>%
  group_by(family) %>%
  summarise("Total Count" = n()) %>%
  kable() %>%
  kable_styling( bootstrap_options = "striped",full_width = F)
#    Family      Total_family
#      0         3405
#      1         3786
```

#### Main Analysis

Figure 1 below displays two boxplots, comparing the average number of pounds of food taken from the UMD food pantry per day between household types (single-member households vs. family households). As expected, those with families have a higher overall mean per day. Additionally, there are more and larger outliers for larger average pounds per day for the family households than single-member households. Similarly, single-member households have more lower outliers for average pounds per day compared to family households. This is expected, since family households have more people to feed than single households.
```{r ggplots1, echo=FALSE}
####################FIG 1######################
#Boxplot of avg pounds of food per household per day, comparing single-member households vs. family households
#As expected, those with families have much higher averages per day, also higher outliers
levels(test1$family) <- c("Single", "Family")

test1 %>% ggplot(aes(x=family, y=mean_food_pounds_per_day, fill=family))+
  geom_boxplot(aes(fill=family))+
  theme_bw() + theme(panel.border = element_blank())+
  labs(x="Household Type",
       y="Average number of pounds of food per day",
       title ="Fig 1: Boxplot of average pounds of food per day by household type")+
  scale_fill_manual(values=c("lightslateblue","steelblue1"))
ggsave("Project1_Fig1_Box.png", height=5, width=7)
```


In Figure 2a below, we see two boxplots of total number of visits in last 10 years per household, comparing single-member households vs. family households. From this figure, we can see that for both household types, the overall average number of total visits in the last 10 years is low. It is also seen from Figure 2a that that there is a larger quantity of outliers outliers for single households than family households. The variance is larger as a result (wider spread). Similarly, some of these outliers are more extreme than those of the family household total visits. One reason for the larger variance in single family household is that single people may come in whenever they need to, but families may try to be more consistent since they may be more aware of the fact that they will need food. This is possibly because of the feeling of higher risk of taking care of someone more than just oneself. Figure 2b compares the boxplots but only for total number of visits that are less than or equal to 60.

```{r ggplots2, echo=FALSE}
####################FIG 2a######################
#Boxplot of total number of visits in last 10 years per household, comparing single-member households vs. family households
#More (and more extreme) outliers for single households, wider spread, than family households
test1 %>% 
  ggplot(aes(x=family, y= Number_of_Visits, fill=family))+
  geom_boxplot(aes(fill=family))+
  theme_bw() + theme(panel.border = element_blank())+
      scale_fill_manual(values=c("lightslateblue","steelblue1"))+
  labs(x="Household Type",
       y="Total number of visits per household",
       title ="Fig 2a: Boxplot of total visits by household type")
ggsave("Project1_Fig2a_Box.png", height=5, width=5)



#Boxplot of total number of visits less than 60 in last 10 years per household, comparing single-member households vs. family households
test1 %>% 
  filter(Number_of_Visits <60) %>% #removing extreme outliers for Number_of_Visits (mostly for single household)
  ggplot(aes(x=family, y= Number_of_Visits, fill=family))+
  geom_boxplot(aes(fill=family))+
  theme_bw() + theme(panel.border = element_blank())+
      scale_fill_manual(values=c("lightslateblue","steelblue1"))+
  labs(x="Household Type",
       y="Total number of visits per household",
       title ="Fig 2b: Boxplot of total visits by household type")
ggsave("Project1_Fig2b_Box.png", height=5, width=5)
```


To examine the data from Figure 2a in a different way, Figure 3a is a histogram comparison of total number of visits between families vs single people. Generally, we can see there are more visits from single households, as well as the single household number of visits distribution having a greater spread. We note that for both households, the largest count of total number of visits is low, close to 1. Figure 3b gives us a closer look at the bulk of the distribution. We truncate our dataset to households with no more than 10 total visits to UMD within our date range, and examine the distributions. We can see more clearly that there are more family than single visits, although there are more long-term visits from single households. As mentioned previously, this could be due to family households trying harder to find homes for their family (rather than just theirselves.)

```{r ggplots3, echo=FALSE}
####################FIG 3a######################
#Histogram comparison of total number of visits between families vs single people.
#More visits from single households, as well as greater spread.
#For both types, largest count is LOW-count visits like 1-3 etc
test1 %>% 
    filter(Number_of_Visits > 10) %>% #removing extreme outliers for Number_of_Visits (mostly for single household)
  ggplot(aes(x=Number_of_Visits))+
  geom_bar(aes(fill=family)) +
    guides(fill=FALSE)+
  scale_fill_manual(values=c("lightslateblue","steelblue1"))+
  theme_bw() + theme(panel.border = element_blank())+
  labs(x="Number of Visits",
       y="Total Households",
       title ="Fig 3a: Distribution of number of visits greater than 10 by household type")+
  facet_wrap(vars(family))
ggsave("Project1_Fig3a_Bar.png", height=4, width=7)
####################FIG 3b######################
test1 %>%
  filter(Number_of_Visits <=10) %>% #removing extreme outliers for Number_of_Visits (mostly for single household)
  ggplot(aes(x=Number_of_Visits, group=family)) +
  geom_bar(position="dodge",aes(fill=family))+
    scale_fill_manual(values=c("lightslateblue","steelblue1"))+
  theme_bw() + theme(panel.border = element_blank())+
  labs(x="Number of Visits",
       y="Total Households",
       title ="Fig 3b: Distribution of number of visits (up to 10) by household type",
       fill="Household Type")
#define family as more than 1 person, as assumed by "provided.food.for"
ggsave("Project1_Fig3b_Bar.png", height=4, width=7)
#More family single visits, more long-term visits from single households
#Could be due to family households try harder to escape homelessness for their kids?
```

In conclusion, we examined two trends comparing household types. The first trend confirms that families have on average greater average pounds of food per day from the pantry. The second trend reveals that families tend to visit less often than single households. However, most households for both household types tend to have "fewer visits" compared to "more visits".

#### References
1) https://www.usich.gov/homelessness-statistics/nc/
2) http://www.umdurham.org/